### 2020-07-20 paper.tex
tweaks

The contributions of this paper include:
\begin{enumerate}
\item A measure-theoretic formulation of involutive MCMC on general state spaces.
\item A mathematical formulation of state spaces consisting of arbitrary key-value stores (i.e. dictionaries) that formalizes the space of execution traces of Gen probabilistic programs, and a formulation of involutive MCMC on these spaces.
\item A differentiable programming language for defining bijections between spaces of traces of probabilistic programs.
\item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
\item An extension to the algorithm that exploits sparsity in the involution to reduce the number of operations needed to apply an involutive MCMC kernel from $O(N^3)$ to $O(1)$ in some cases, where $N$ is the dimensionality of the latent space.
\item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
\end{enumerate}

@@

The involutive MCMC construction encompasses many existing classes of MCMC kernels, some of which explicitly make use of bijective or involutive deterministic maps.
In particular, the reversible jump framework~\citep{green1995reversible,hastie2012model} employs a family of continuously differentiable bijections between the parameter spaces of different models.
\citet{tierney1998note} described a family of deterministic proposals based on a deterministic involution that is equivalent to involutive MCMC but without the auxiliary probability distribution.
More recently,~\citet{spanbauer2020deep} defined a class of deep generative model based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.

@@

    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}_b, x \oplus y)$

@@

\Procedure{run-involution}{$\mathcal{F}_b$, $z$}
% TODO would be much simpler to describe if we didn't have to separate out x and y..

@@

@@

Our probabilistic programming language augments the syntax of Julia~\citep{bezanson2017julia}
with a single new construct, the `$\{\mathtt{address}\} \sim \mathtt{distribution}$' expression, for making a \textit{named} random choice. An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form `$\{\mathtt{address}\} \sim \mathtt{distribution}$', (i) evaluating the \textit{address expression} (`$\mathtt{address}$') to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} (`$\mathtt{distribution}$') to obtain a probability distribution over the measurable space $V_k$; (iii) sampling a value $x_k$ from this distribution and adding the mapping $\{k \mapsto x_k\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution. When execution terminates, the program trace has accumulated a mapping for each random choice encountered during the program's execution.

@@

Typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC). Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses. Let $\tilde{p}$ denote $\mathcal{P}$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}. Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{\mathcal{L}(b)}$, where $\mathcal{L}(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the marginal likelihood of $b$, and does not depend on $x$. Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$. If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio of densities in Algorithm~\ref{alg:involutive-mcmc}, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
= \frac{\tilde{p}(x' \oplus b) q_{x'}(y')}{\tilde{p}(x \oplus b) q_x(y)}
= \frac{\tilde{p}(x' | b) q_{x'}(y')}{\tilde{p}(x | b) q_x(y)}
\end{equation}

@@

Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution (and the resulting cancellations in the acceptance ratio) for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

@@

Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then the density ratio in the acceptance probability for involutive MCMC on dictionaries (Equation~(\ref{eq:acceptance-ratio-dictionaries})) can be automated using probabilistic programming techniques.
This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.

@@

We now introduce a simple differentiable programming language for specifying involutions $f$.
The language needs to have syntax for reading the value from an address in $x \oplus y \in Z$ and writing to an address in $x' \oplus y' \in Z$.
To read a value $(x \oplus y)[a]$ at address $a$, we use the \texttt{@read} keyword:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
    value = @read(<address>, <type>)
\end{lstlisting}}
\end{tabular}
\end{center}
The first argument is the address $a$ and the second argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the interpreter whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
Recall that $x$ is the trace of the model probabilistic program, and $y$ is the trace of the auxiliary probabilistic program.
Each address $a$ therefore needs to specify which of these traces to read from, and the address within that trace.
The traces are given names in the function signature:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@transform f (model_in, aux_in) to (model_out, aux_out)
begin
..
end
\end{lstlisting}
\end{lrbox}%
\scalebox{0.94}{\usebox{\mybox}}
\end{tabular}
\end{center}
Here, the traces $x$, $y$, $x'$ and $y'$ are given names $\mathtt{model\_in}$, $\mathtt{aux\_in}$, $\mathtt{model\_out}$, and $\mathtt{aux\_out}$, respectively.
The syntax for address $\mathtt{a}$ within trace $\mathtt{trace}$ is $\mathtt{trace[a]}$.
For example, to read the value of a continuous address $\mathtt{:}\mathtt{r}$ from the input model trace ($x$):
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
    val = @read(model_in[:r], :continuous)
\end{lstlisting}}
\end{tabular}
\end{center}
The syntax for writing to an address in $x' \oplus y' \in Z$ is similar.
For example to write a value $\mathtt{val}$ to $x'[\mathtt{:}\mathtt{r}]$:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
    @write(model_in[:r], val, :continuous)
\end{lstlisting}}
\end{tabular}
\end{center}
Note that the input traces $x, y$ are distinct from the output traces $x', y'$;
input traces can only be read from, and output traces can only be written to.
For example, it is not possible to write an output trace and then read the written value from it later.

Often, we want to simply copy the value from some address in the input traces to some address in the output trace.
While this is possible via a $\mathtt{@read}$ followed by a $\mathtt{@write}$, the language provides a special syntax:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
    @copy(<from-address>, <to-address>)
\end{lstlisting}}
\end{tabular}
\end{center}
For example, to copy the value from address $\mathtt{:}\mathtt{u}$ in $x$ to address $\mathtt{:}\mathtt{v}$ in $x'$, we use:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
    @copy(model_in[:u], model_in[:v])
\end{lstlisting}}
\end{tabular}
\end{center}
Of course, it is also possible to copy from $x$ to $y'$, from $y$ to $x'$ and from $y$ to $y'$.
As we will see in Section~\ref{sec:sparsity}, it is preferable to use $\mathtt{@copy}$ when possible instead of reading and then writing, as this can improve the efficiency of automatic differentiation.

%We denote a program in the differentiable language by $\mathcal{F}$ and the involution it defines as $f$.
%Like the probabilistic programming language described in the previous section, our differentiable language for involutions on traces is embedded based on Julia.
%The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
%The first argument defines what trace we are reading from.
%The second argument is the address in that trace, which may be hierarchical.\footnote{
%Hierarchical addresses of the form $\mathtt{a.b.c}$ are constructed in the language using the syntax \texttt{a=>b=>c}.}
%The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
%Similarly, to write a value to a trace, we use a \texttt{@write} keyword:
%\noindent
%\begin{center}
%\begin{tabular}{c}
%{\begin{lstlisting}[basicstyle=\small\ttfamily]
    %@write(<trace>, <address>, <value>, <type>)
%\end{lstlisting}}
%\end{tabular}
%\end{center}
%The mathematical behavior of the function is defined simply by writing Julia code that implements it.
\paragraph{Constructing an Involution}
Consider the following generative model, which posits that a vector of univariate data is either generated from a single normal distribution or a mixture of two normal distributions.
The model is expressed as a probabilistic program:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function p()
  k ~ uniform_discrete(1, 2)
  means = [{(:mu, j)} ~ normal(0, 10) for j in 1:k]
  weights = ones(k)/k
  vars = ones(k)
  for i in 1:100
    {(:x, i)} ~ mixture_of_normals(weights, means, vars)
  end
end
\end{lstlisting}
\end{lrbox}%
\scalebox{0.90}{\usebox{\mybox}}
\end{tabular}
\end{center}
This is a simplified version of the unfinite Gaussian mixture model in Figure~\ref{fig:mixture}b.
Each key of the form $(\mathtt{:}\mathtt{x}, i)$ will be observed (in the dictionary $b$), and the latent part of the trace (the dictionary $x$) contains the other keys $\mathtt{:}\mathtt{k}$ and $(\mathtt{:}\mathtt{mu}, 1)$ and (only if $\mathtt{k} = 2$) $(\mathtt{:}\mathtt{mu}, 2)$.

We will now walk through how to express a simple `split-merge' move~\citep{richardson1997bayesian} using our differentiable programming language.
Suppose we want to construct an involutive MCMC kernel that changes $\mathtt{k}$ from $2$ to $1$ or vice-versa.
The discrete part of the involution is straightforward:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
k = @read(model_in["k"], discrete)
if k == 1
    @write(model_out["k"], 2, discrete)
else
    @write(model_out["k"], 1, discrete)
end
\end{lstlisting}
\end{lrbox}%
\scalebox{1.0}{\usebox{\mybox}}
\end{tabular}
\end{center}
Writing the involution code for the continuous choices is more complex.
There is no one-to-one correspondence between the space of dictionaries with keys $\{\mathtt{:}\mathtt{k}, (\mathtt{:}\mathtt{mu}, 1)\}$ and the space of dictionaries $x$ with keys $\{\mathtt{:}\mathtt{k}, (\mathtt{:}\mathtt{mu}, 1), (\mathtt{:}\mathtt{mu}, 2)\}$.
Therefore, we need to extend the state space using the auxiliary distribution, defined with the following probabilistic program:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function q(model_trace)
    if model_trace["k"] == 1
        # we are doing a split, sample extra DoF
        u ~ beta(2, 2)
    end
end
\end{lstlisting}
\end{lrbox}%
\scalebox{1.00}{\usebox{\mybox}}
\end{tabular}
\end{center}
Our involution is on the space of combined dictionaries $x \oplus y$ where $y$ are traces of $\mathtt{q}$.
Note that for each $(x \oplus y)$ where $x[\mathtt{k}] = 1$, $y$ has a key $\mathtt{u}$, and for each $(x \oplus y)$ where $x[\mathtt{k}] = 2$, $y$ is empty.
We extend the involution using a pair of bijections between $(\mu_1, \mu_2)$ and $(\mu_1, u)$:
\[
\mu_1, \mu_2 \mapsto ((\mu_1 + \mu_2) / 2, \mu_2 - (\mu_1 + \mu_2)/2)
\]
\[
\mu_1, u \mapsto (\mu_1 - u, \mu_1 + u)
\]
The value of $x[\mathtt{k}]$ determines which of these functions is executed.
The full involution program $\mathcal{F}$ is then:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@transform f (model_in,aux_in) to (model_out,aux_out)
begin
    k = @read(model_in["k"], discrete)
    if k == 1
        @write(model_out["k"], 2, discrete)
        mu = @read(model_in[("mu", 1)], continuous)
        u = @read(aux_in["u"], continuous)
        mu1 = mu - u; mu2 = mu + u
        @write(model_out[("mu", 1)], mu1, continuous)
        @write(model_out[("mu", 2)], mu2, continuous)
    else
        @write(model_out["k"], 1, discrete)
        mu1 = @read(model_in[("mu", 1)], continuous)
        mu2 = @read(model_in[("mu", 2)], continuous)
        mu = (mu2 + mu1) / 2
        u = mu2 - mu
        @write(model_out[("mu", 1)], mu, continuous)
        @write(aux_out["u"], u, continuous)
    end
end
\end{lstlisting}
\end{lrbox}%
\scalebox{0.90}{\usebox{\mybox}}
\end{tabular}
\end{center}
The two continuous bijections (implemented in blocks of code in the two branches) are inverses of one another.
This is a common pattern in involution programs---an involution on the discrete parts of the traces (in this case, just $x[\mathtt{k}]$) determines via control flow which continuous code blocks get executed, such that the end-to-end program defines an involution.

Note that the observations $b$ (in this case, the x-coordinates) are included in traces of $\mathtt{p}$, which take the form $x \oplus b$ where $x$ is the latent part and $b$ is the observed part.
The involution program $\mathcal{F}$ is allowed to read from the the observations in the model trace using the same syntax as used to read from the latent part of the trace $x$.
To denote this dependency we use notation $\mathcal{F}_b$.
(The example given here does not utilize this feature).

@@

Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[k]$ for $k \in K_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[k]$ for $k \in K_{z'} \cap \mathcal{J}$).
We compute the Jacobian using automatic differentiation. (The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.)
The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD, this is accomplished by iterating over output continuous addresses and backpropagating from each one to all of the input continuous addresses, computing the Jacobian column-by-column.
Keys with an $n$-dimensional Lebesgue measure as their reference measure, corresponding to vector-valued random choices, are unpacked into $n$ separate columns in the Jacobian; this detail is elided in Algorithm~\ref{alg:auto-involutive-mcmc}.
The $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are also omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).

>>>>

The contributions of this paper include:
\begin{enumerate}
\item A measure-theoretic formulation of involutive MCMC on general state spaces.
\item A mathematical formulation of state spaces consisting of arbitrary key-value stores (i.e. dictionaries) that formalizes the space of execution traces of Gen probabilistic programs, and a formulation of involutive MCMC on these spaces.
\item A differentiable programming language for defining transformations between spaces of traces of probabilistic programs.
\item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
\item An extension to the algorithm that exploits sparsity in the involution to reduce the number of operations needed to apply an involutive MCMC kernel from $O(N^3)$ to $O(1)$ in some cases, where $N$ is the dimensionality of the latent space.
\item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
\end{enumerate}

@@

The involutive MCMC construction encompasses many existing classes of MCMC kernels, some of which explicitly make use of bijective or involutive deterministic maps.
In particular, the reversible jump framework~\citep{green1995reversible,hastie2012model} employs a family of continuously differentiable bijections between the parameter spaces of different models.
\citet{tierney1998note} described a family of deterministic proposals based on a deterministic involution that is equivalent to involutive MCMC but without the auxiliary probability distribution.
More recently,~\citet{spanbauer2020deep} defined a class of deep generative models based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.

@@

    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y, b)$

@@

\Procedure{run-involution}{$\mathcal{F}$, $z$, $b$}

@@

        \State (and reading from observations $b$ for any $k \in K_b$)

@@

Our probabilistic programming language augments the syntax of Julia~\citep{bezanson2017julia}
with a single new construct, the `$\{\mathtt{address}\} \sim \mathtt{distribution}$' expression, for making a \textit{named} random choice. An execution trace of a Gen probabilistic program is a dictionary that can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form `$\{\mathtt{address}\} \sim \mathtt{distribution}$', (i) evaluating the \textit{address expression} (`$\mathtt{address}$') to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} (`$\mathtt{distribution}$') to obtain a probability distribution over the measurable space $V_k$; (iii) sampling a value $x_k$ from this distribution and adding the mapping $\{k \mapsto x_k\}$ to the execution trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution. When execution terminates, the execution trace has accumulated a mapping for each random choice encountered during the program's execution.

@@

Typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC). Instead, we may wish to sample from a target $p$ that arises from \emph{conditioning} a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses. Let $\tilde{p}$ denote $\mathcal{P}$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}. Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{\mathcal{L}(b)}$, where $\mathcal{L}(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the marginal likelihood of $b$, and does not depend on $x$. Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$. If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio of densities in Algorithm~\ref{alg:involutive-mcmc}, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
= \frac{\tilde{p}(x' \oplus b) q_{x'}(y')}{\tilde{p}(x \oplus b) q_x(y)}
= \frac{\tilde{p}(x' | b) q_{x'}(y')}{\tilde{p}(x | b) q_x(y)}
\end{equation}

@@

Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and cancellations in the acceptance ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

@@

Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then the density ratio in the acceptance probability for involutive MCMC on dictionaries (Equation~(\ref{eq:acceptance-ratio-dictionaries})) can be automated using probabilistic programming techniques.
This section shows that if the involution $f$ is specified using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.

@@

We now introduce a simple differentiable programming language for specifying involutions $f$.
The language needs to have syntax for reading the value from an address in $x \oplus y \in Z$ and writing to an address in $x' \oplus y' \in Z$.
To read a value $(x \oplus y)[a]$ at address $a$, we use the \texttt{@read} keyword:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
    value = @read(<address>, <type>)
\end{lstlisting}}
\end{tabular}
\end{center}
The first argument is the address $a$ and the second argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the interpreter whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
Recall that $x$ is the trace of the model probabilistic program, and $y$ is the trace of the auxiliary probabilistic program.
Each address $a$ therefore needs to specify which of these traces to read from, and the address within that trace.
The traces are given names in the function signature:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@transform f (model_in, aux_in) to (model_out, aux_out)
begin
..
end
\end{lstlisting}
\end{lrbox}%
\scalebox{0.94}{\usebox{\mybox}}
\end{tabular}
\end{center}
Here, the traces $x$, $y$, $x'$ and $y'$ are given names $\mathtt{model\_in}$, $\mathtt{aux\_in}$, $\mathtt{model\_out}$, and $\mathtt{aux\_out}$, respectively.
The syntax for address $\mathtt{a}$ within trace $\mathtt{trace}$ is $\mathtt{trace[a]}$.
For example, to read the value of a continuous address $\mathtt{:}\mathtt{a}$ from the input model trace ($x$):
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
val = @read(model_in[:a], :continuous)
\end{lstlisting}}
\end{tabular}
\end{center}
The syntax for writing to an address in $x' \oplus y' \in Z$ is similar.
For example to write a value $\mathtt{val}$ to $x'[\mathtt{:}\mathtt{a}]$:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@write(model_out[:a], val, :continuous)
\end{lstlisting}}
\end{tabular}
\end{center}
Note that the input traces $x, y$ are distinct from the output traces $x', y'$;
input traces can only be read from, and output traces can only be written to.
For example, it is not possible to write an output trace and then read the written value from it later.

Often, we want to simply copy the value from some address in the input traces to some address in the output trace.
While this is possible via a $\mathtt{@read}$ followed by a $\mathtt{@write}$, the language provides a special syntax:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(<source-address>, <destination-address>)
\end{lstlisting}}
\end{tabular}
\end{center}
For example, to copy the value from address $\mathtt{:}\mathtt{u}$ in $x$ to address $\mathtt{:}\mathtt{v}$ in $x'$, we use:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@copy(model_in[:u], model_out[:v])
\end{lstlisting}}
\end{tabular}
\end{center}
Of course, it is also possible to copy from $x$ to $y'$, from $y$ to $x'$ and from $y$ to $y'$.
As we will see in Section~\ref{sec:sparsity}, it is preferable to use $\mathtt{@copy}$ when possible instead of reading and then writing, as this can make the acceptance probability calculation more efficient.

%We denote a program in the differentiable language by $\mathcal{F}$ and the involution it defines as $f$.
%Like the probabilistic programming language described in the previous section, our differentiable language for involutions on traces is embedded based on Julia.
%The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
%The first argument defines what trace we are reading from.
%The second argument is the address in that trace, which may be hierarchical.\footnote{
%Hierarchical addresses of the form $\mathtt{a.b.c}$ are constructed in the language using the syntax \texttt{a=>b=>c}.}
%The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
%Similarly, to write a value to a trace, we use a \texttt{@write} keyword:
%\noindent
%\begin{center}
%\begin{tabular}{c}
%{\begin{lstlisting}[basicstyle=\small\ttfamily]
    %@write(<trace>, <address>, <value>, <type>)
%\end{lstlisting}}
%\end{tabular}
%\end{center}
%The mathematical behavior of the function is defined simply by writing Julia code that implements it.
\paragraph{Constructing an Involution}
Consider the following generative model, which posits that a vector of univariate data is either generated from a single normal distribution or a mixture of two normal distributions.
The model is expressed as a probabilistic program:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function p()
  k ~ uniform_discrete(1, 2)
  means = [{(:mu, j)} ~ normal(0, 10) for j in 1:k]
  weights = ones(k)/k
  vars = ones(k)
  for i in 1:100
    {(:x, i)} ~ mixture_of_normals(weights, means, vars)
  end
end
\end{lstlisting}
\end{lrbox}%
\scalebox{0.90}{\usebox{\mybox}}
\end{tabular}
\end{center}
This is a simplified version of the infinite Gaussian mixture model in Figure~\ref{fig:mixture}b.
Each key of the form $(\mathtt{:}\mathtt{x}, i)$ will be observed (in the dictionary $b$), so the latent part of the trace (the dictionary $x$) contains the other keys $\mathtt{:}\mathtt{k}$ and $(\mathtt{:}\mathtt{mu}, 1)$ and (when $\mathtt{k} = 2$) $(\mathtt{:}\mathtt{mu}, 2)$.

We will now walk through how to express a simple `split-merge' move~\citep{richardson1997bayesian} using our differentiable programming language.
Suppose we want to construct an involutive MCMC kernel that changes $\mathtt{k}$ from $2$ to $1$ or vice versa.
The discrete part of the involution is straightforward:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
k = @read(model_in["k"], discrete)
if k == 1
    @write(model_out["k"], 2, discrete)
else
    @write(model_out["k"], 1, discrete)
end
\end{lstlisting}
\end{lrbox}%
\scalebox{1.0}{\usebox{\mybox}}
\end{tabular}
\end{center}
Writing the involution code for the continuous choices is more complex.
There is no one-to-one correspondence between the space of dictionaries with keys $\{\mathtt{:}\mathtt{k}, (\mathtt{:}\mathtt{mu}, 1)\}$ and the space of dictionaries $x$ with keys $\{\mathtt{:}\mathtt{k}, (\mathtt{:}\mathtt{mu}, 1), (\mathtt{:}\mathtt{mu}, 2)\}$.
Therefore, we need to extend the state space using the auxiliary distribution, defined with the following probabilistic program:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function q(model_trace)
    if model_trace["k"] == 1
        # we are doing a split, sample extra DoF
        u ~ beta(2, 2)
    end
end
\end{lstlisting}
\end{lrbox}%
\scalebox{1.00}{\usebox{\mybox}}
\end{tabular}
\end{center}
Our involution is on the space of combined dictionaries $x \oplus y$ where $y$ are traces of $\mathtt{q}$.
Note that for each $(x \oplus y)$ where $x[\mathtt{k}] = 1$, $y$ has a key $\mathtt{u}$, and for each $(x \oplus y)$ where $x[\mathtt{k}] = 2$, $y$ is empty.
We extend the involution using a pair of bijections between $(\mu_1, \mu_2)$ and $(\mu_1, u)$ that show how two cluster means should be transformed into one cluster mean (merge), and vice versa (split):
\begin{equation} \label{eq:split-merge}
\begin{array}{c}
\mu_1, \mu_2 \mapsto ((\mu_1 + \mu_2) / 2, \mu_2 - (\mu_1 + \mu_2)/2) \;\; [\mathrm{Merge}]\\
\mu_1, u \mapsto (\mu_1 - u, \mu_1 + u) \;\; [\mathrm{Split}]
\end{array}
\end{equation}
The value of $x[\mathtt{k}]$ determines which of these functions is executed.
The full involution program $\mathcal{F}$ is then:
\noindent
\begin{center}
\begin{tabular}{c}
\begin{lrbox}{\mybox}%
\begin{lstlisting}[basicstyle=\small\ttfamily]
@transform f (model_in,aux_in) to (model_out,aux_out)
begin
    k = @read(model_in["k"], discrete)
    if k == 1
        @write(model_out["k"], 2, discrete)

        # split bijection
        mu = @read(model_in[("mu", 1)], continuous)
        u = @read(aux_in["u"], continuous)
        mu1 = mu - u; mu2 = mu + u
        @write(model_out[("mu", 1)], mu1, continuous)
        @write(model_out[("mu", 2)], mu2, continuous)
    else
        @write(model_out["k"], 1, discrete)

        # merge bijection
        mu1 = @read(model_in[("mu", 1)], continuous)
        mu2 = @read(model_in[("mu", 2)], continuous)
        mu = (mu2 + mu1) / 2
        u = mu2 - mu
        @write(model_out[("mu", 1)], mu, continuous)
        @write(aux_out["u"], u, continuous)
    end
end
\end{lstlisting}
\end{lrbox}%
\scalebox{0.90}{\usebox{\mybox}}
\end{tabular}
\end{center}
The two continuous bijections (implemented in blocks of code in the two branches) are inverses of one another.
This is a common pattern in involution programs---an involution on the discrete parts of the traces (in this case, just $x[\mathtt{k}]$) determines via control flow which continuous code blocks get executed, such that the end-to-end program defines an involution.

Note that the observations $b$ (in this case, the x-coordinates) are included in traces of $\mathtt{p}$, which take the form $x \oplus b$ where $x$ is the latent part and $b$ is the observed part.
The involution program $\mathcal{F}$ is allowed to read from the the observations in the model trace using the same syntax as used to read from the latent part of the trace $x$.
(The example given here does not utilize this feature).

@@

Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[k]$ for $k \in K_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[k]$ for $k \in K_{z'} \cap \mathcal{J}$).
In the example above, the partition $E$ is given by the equivalence classes of the relation $(x_1, y_1) \sim (x_2, y_2) \iff x_1[\mathtt{k}] = x_2[\mathtt{k}]$ (there are two equivalence classes), $g$ maps the $k = 1$ class to the $k = 2$ class and vice versa, and 
the continuous bijections $h_e$ and $h_{g(e)}$ are the functions in Equation~(\ref{eq:split-merge}).
